### 2013-05-19 draft.tex
fixed 1st 5 paragraphs of constraining temp section: added solid-state, 35 degrees brachs, 5% estimates

Independent assessments of preservation quality confirm that the lowest-temperature brachiopods are very well preserved. To provide additional support for this hypothesis, we measured the clumped isotope composition of the micrite adjecent to four brachiopods (Figure \ref{comparisons}). In all four cases, the temperature of the micrite was resolvably higher than the temperature of the brachiopod. We expect micrite to be less pristine than our brachiopod samples because micrite is fine-grained, not protected by a layer of brachiopod shell, and was not picked under a microscope prior to analysis. Because the less pristine material, micrite, was associated with higher temperatures, these comparisons are consistent with our hypothesis that the lowest temperature samples are the most pristine.

The analyses described above have allowed us to define the temperature consequences of diagenesis, but to confidently interpret the \deltao O of Silurian seawater, we also need to characterize the oxygen isotopic consequences of diagenesis. To constrain the effect of diagenetic alteration on the \deltao O of water calculated from the samples, we plotted the temperature versus the \deltao O of water for all of the samples in this study (Figure \ref{ellipses}). There is a clear positive relationship between the temperature and \deltao O of water, indicating that interstitial water grew increasingly enriched in $^{18}$O as the temperature increased. This relationship is diagnostic of diagenesis occurring in the presence of a low water-to-rock ratio. In this diagenetic scheme, also called rock-buffered diagenesis, as the temperature increases, the equilibrium oxygen isotope fractionation between water and carbonate grows smaller, preferentially driving $^{18}$O into the water. Importantly, rock-buffered diagenesis causes the temperature of more altered samples to increase, which is consistent with our hypothesis that the lowest temperature samples are the most pristine.

The lowest, most-pristine sample in this study (Figure \ref{ellipses}) indicates that tropical Silurian oceans had a temperature of $30\pm 2$\degrees C and a \deltao O of $-1.7\pm 0.5\permil$. One caveat of these values is that they are based on only one measurement (G13 10.5 to 10.8 B1). The first time this brachiopod was picked, it produced enough material for two analyses, but one analysis was rejected because its $\Delta_{48}$ was more than 1$\permil$ above the heated gas standard line. Later, more material was picked from the same bachiopod, but in order to get enough material for a replicate measurement, very small flakes had to be included. We worried that the small flakes would be lost in the rinsing procedure, so this replicate sample was not rinsed. The temperature of the replicate sample (G13 10.5 to 10.8 B1 leftovers) was $41\pm 4$ degrees C, which may reflect contamination that ordinarily would have been removed during the rinse step of sample preparation. Due to the non-standard treatment of this replicate sample, we do not treat it as a true replicate, and we report its temperature separately.

A second caveat of our estimates of the temperature and \deltao O of Silurian tropical oceans is that they are technically upper estimates. We are confident that our sample suite was subjected to a diagenetic regime that increased the temperature and \deltao O of water recorded by the samples, and we are not able to exclude the possibility that diagenesis affected all the samples in our study, even the most pristine sample. However, the micro-scale studies of the calcite fabric of the lowest-temperature brachiopods indicate that late-phase calcite is only a trace constituent of these samples, suggesting that the upper estimate is likely very similar to the true value.

@@

Our results, along with those of earlier clumped isotope studies \citep{Came2007,Finnegan2011,Dennis2013}, provide consistent explanations for the low \deltao O values of early Paleozoic carbonates. Carbonate clumped isotopes constrain the contribution of a change in seawater \deltao O to only a few permil of the observed trend, ruling out a value of $-8\permil$ for the \deltao O of early Paleozoic oceans \citep{Grossman2012}. Our results also serve as a reminder that many of the fossils preserved in the geologic record lived in tropical oceans. The \deltao O of carbonate in equilibrium with tropical ocean water should not be compared to the \deltao O of carbonate in equilibrium with water of average modern sea surface temperature, because tropical oceans are much warmer than the average sea surface temperature. Clumped isotopes additionally demonstrate that extremely hot ($\sim 70$\degrees C) ocean temperatures cannot be the cause of the depleted \deltao O values of early Paleozoic carbonates. Finally, clumped isotopes, as a highly sensitive indicator of diagenetic alteration, demonstrate that even among samples with original calcite fabric and pristine trace metal signatures, there is a large range in preservation quality. This means that the average fossil calcite in almost any sample suite is likely altered from its primary composition, and that diagenetic alteration is a key contributor to the low \deltao O values of early Paleozoic carbonates.

>>>>

% comparisons support hypothesis that low temp most pristine
% array shows rock-buffered diagenesis
% most pristine sample only one replicate
% only upper estimates because all samples could have a little contamination
% Silurian warm and similar d18O to today
% resolve veizer controversy
% surprising that d18O has not changed

% array. probably not solid-state reordering, so we interpret rock-buffered diagenesis
% comparisons support hypothesis that low temp most pristine, EBSD supports, metals dont negate
% most pristine samples population at 34, one at 30, resample not replicate, no big deal!
% only upper estimates, but probably pretty close
% Silurian warm and similar in d18O to today. matches previous clumped studies
% resolves veizer controversy
% implications of warm oceans for Ireviken climate rule out big changes, can't resolve small
% implications of constant d18O?? relative amounts constant. either both constant or coupled.

To estimate the temperature and \deltao O of Silurian oceans, we first need to confirm the hypothesis that the lowest-temperature samples are the most pristine and therefore record the most reliable record of Silurian ocean conditions. This hypothesis can be tested through constraints on the diagenetic regime to which the samples were subjected. As outlined above, different diagenetic regimes have characteristic effects on the relationship between temperature and the \deltao O of interstitial water. When the clumped isotope temperatures of the samples in this study are plotted versus the \deltao O of water calculated for each sample, there is a clear positive relationship (Figure \ref{ellipses}). This relationship suggests that interstitial water grew increasingly enriched in $^{18}$O as the temperature increased, which is diagnostic of rock-buffered diagenesis. Importantly, during rock-buffered diagenesis, more altered samples have higher temperatures, which is consistent with our hypothesis that the lowest temperature samples are the most pristine.

Solid-state diffusion is commonly considered as a possible mechanism for the increase of clumped isotope temperatures during diagenesis, but we rule it out as a major contributor to our results. During solid-state diffusion, isotopes of carbon and oxygen diffuse through the calcite mineral lattice, which decreases the degree of clumping and increases the interpreted temperature of the calcite. This process does not affect the bulk isotopic composition of the calcite, and it occurs without dissolution and reprecipitation, which means it is also not expected to change the calcite fabric or trace metal content \citep{Eiler2011}. However, solid-state diffusion is expected to affect all calcite approximately equally, which means that this process cannot explain the variation in temperature among the samples. Additionally, solid-state diffusion is thought to occur very slowly at temperatures below 100\degrees C (D. Stolper personal communication 2013), which is the maximum temperature experienced by Gotland strata, as constrained by the conodont alteration index \citep{Jeppsson1983,Wenzel2000}. Solid-state diffusion is not consistent with our observations, leaving rock-buffered diagenesis as the most likely characterization of the diagenetic regime.

Comparisons of the clumped isotope temperature of different calcite textures provide additional support for the hypothesis that the samples underwent rock-buffered diagenesis. We measured the clumped isotope composition of the micrite adjecent to four brachiopods, and in all four cases, the temperature of the micrite was resolvably higher than the temperature of the brachiopod (Figure \ref{comparisons}). Micrite is expected to be less pristine than the brachiopod samples because micrite is fine-grained, not protected by a layer of brachiopod shell, and was not picked under a microscope prior to analysis. Because the less pristine material, micrite, was associated with higher temperatures, these comparisons are consistent with the hypothesis that the lowest temperature samples are the most pristine.

The lowest-temperature, most pristine sample in this study, a brachiopod, indicates that tropical Silurian oceans had a temperature of $30\pm 2$\degrees C and a \deltao O$_{\text{SMOW}}$ of $-1.7\pm 0.5\permil$ (Figure \ref{ellipses}), where the error estimates represent the standard error of the mean over eight acquisitions. The two next-most pristine samples, also brachiopods, have temperatures of $33\pm 2$\degrees C and $35\pm 2$\degrees C, which correspond to seawater \deltao O$_{\text{SMOW}}$ values of $-1.5\pm 0.3\permil$ and $-1.9\pm 0.4\permil$, respectively. One caveat of these values is that they are each based on only one measurement due to sample-size limitations. The lowest-temperature brachiopod was picked a second time, but to collect enough material for analysis, some less pristine flakes had to be included, which may explain the $41\pm 4$\degrees C temperature of this re-sample (G13 10.5 to 10.8 B1 leftovers). However, we argue that the differences between the temperatures and \deltao O values implied by the three most pristine samples are small enough that they essentially represent the same warm, tropical Silurian ocean water.

A second caveat of our estimates of the temperature and \deltao O of Silurian tropical oceans is that they are technically upper estimates. We are confident that our sample suite was subjected to a diagenetic regime that increased the temperature and \deltao O of water recorded by the samples. We are not able to exclude the possibility that diagenesis affected all the samples in our study, even the most pristine samples. However, micro-scale studies of the calcite fabric of the lowest-temperature brachiopods indicate that late-phase calcite is only a trace constituent of these samples, comprising a maximum of 5\% of the material. Assuming linear mixing, which is not technically correct but close enough for our purposes, if 5\% of the sample material was recrystallized at 100\degrees C and the rest of the sample material retained an original temperature of 30\degrees C, the measured temperature of the sample would be $\sim34$\degrees C. Therefore, the constraints provided by the EBSD mapping of the bachiopod shell calcite demonstrate that our estimate of the temperature of the tropical Silurian ocean is likely reliable to within 4\degrees C.

@@

Our results, along with those of earlier clumped isotope studies \citep{Came2007,Finnegan2011,Dennis2013}, provide consistent explanations for the low \deltao O values of early Paleozoic carbonates. Carbonate clumped isotopes constrain the contribution of a change in seawater \deltao O to only a few permil of the observed trend, ruling out a value of $-8\permil$ \citep{Jaffres2007,Veizer1999} for the \deltao O of early Paleozoic oceans \citep{Grossman2012}. Our results also serve as a reminder that many of the fossils preserved in the geologic record lived in tropical oceans. The \deltao O of carbonate in equilibrium with tropical ocean water should not be compared to the \deltao O of carbonate in equilibrium with water of average modern sea surface temperature, because tropical oceans are much warmer than the average sea surface temperature. Clumped isotopes additionally demonstrate that extremely hot ($\sim 70$\degrees C) ocean temperatures cannot be the cause of the depleted \deltao O values of early Paleozoic carbonates. Finally, clumped isotopes, as a highly sensitive indicator of diagenetic alteration, demonstrate that even among samples with original calcite fabric and pristine trace metal signatures, there is a large range in preservation quality. This means that the average fossil calcite in almost any sample suite is likely altered from its primary composition, and that diagenetic alteration is a key contributor to the low \deltao O values of early Paleozoic carbonates.
